Master Thesis

### 2013-06-17 implementation.tex
Last night shift

\chapter{Wide-coverage abstract categorial grammars}

@@

Given this setup, there are several approaches we might try, none of
them too appealing. We might write our metagrammar in a different format
and use a tool which would combine the metagrammar and the lexical
database\footnote{We will be using the term \emph{lexical database} to
  refer to what is usually called a lexicon, in order not to cause
  confusion with the lexicons of ACGs.} to produce the lexicalized ACG
in a direct representation, ready to be loaded by the toolkit. This
workflow would make development on the grammar tedious as every time we
would change the metagrammar, we would have to regenerate the direct
representation and then load it in the ACG toolkit. With extra effort,
the former cost could be alleviated by devising a system for
regenerating only parts of the grammars which will have changed since
last time. However, we would still have to pay the price of loading the
entire grammar into the toolkit before being able to interact with it
again.

@@

In this chapter, we will spend time explaining the decisions which went
into designing the system and the form it has now.

\section{The tools and techniques}

@@

By insisting on purity, I/O becomes more difficult as we would not allow
the side-effectful pseudorelations of Prolog. Therefore, we will not
force ourselves to implement the entire program front-to-back using only
relational programming. Instead, we will use a relational programming
system, miniKanren, which is embedded inside a functional programming
language. This lets us reap the benefits of relational programming
inside the core logic of our program while deferring to the functional
programming language to provide I/O and other bookkeeping operations.

\subsection{The case for relational programming}

There are several reasons why we might want to choose a relational
programming system for our implementation. First off, the problems which
consiture the algorithmic core of our program are often mere textbook
examples of how to use a relational programming system. This goes for
type inference/checking/inhabitation, which are all implemented by a
single relation which is often used as a kind of ``Hello World'' program
in the world of relational programming. Similarly, implementing
$\beta$-reduction cleanly (without having to handle variable clashes by
generating new variables and $\alpha$-converting) is a motivational
example for introducing \emph{nominal logic programming}. To reach
feature parity with the ACG toolkit, one would only need to implement
the mapping of a lexicon over a term, which is a simple functor.

@@

Finally, some of the structures that we will encounter in our domain are
more directly modelled in mathematics as a relation instead of a
function. This has the benefit that to implement a relation in
relational programming, one just writes the relation, while if we were
forced to encode everything in functions, we would have to manage this
multiplicity of values manually. Examples of structures that have this
kind of structures if Frilex, our lexical database, where a single
wordform can be mapped to multiple hypertags. We will see another
example of a relation within our domain in the next section and we
will also discuss the advantages and drawbacks of modelling functions as
relations.

\subsection{Choice of implementation language}

There are two languages which play host to a substantial implementation
of miniKanren (meaning one that has extensible unification and/or
constraints and that includes support for nominal logic),
Racket\footnote{\url{https://github.com/calvis/cKanren}} and
Clojure\footnote{\url{https://github.com/clojure/core.logic}}. The two
languages are very similar to each other and therefore the choice is of
no great significance. An appealing feature of Clojure is that it uses
abstract collection types pervasively instead of tying the standard
operations to specific structures such as lists and cons cells and that
it follows other similarly enlightened design decisions. Racket on the
other hand has had more time to mature, has better metaprogramming
facilities and is more recognized in academia. In developing our system,
we have settled for Clojure. One of the advantages of doing so is that
the Clojure implementation of miniKanren, dubbed core.logic, has
inspired the software community and the project has attracted
contributors that submit issue reports and
patches\footnote{\url{http://dev.clojure.org/jira/browse/LOGIC}}.

\section{Defining signatures and lexicons}

\subsection{Defining signatures}

In~\ref{ssec:sig}, we formally presented signatures as triples
$\mathopen{<}A, C, \tau\mathclose{>}$. The set $C$ can be easily
reconstructed as the domain of $\tau$ and the set $A$ is just the set of
all atomic types occurring in the range of $\tau$. Therefore, to define
the signature $\mathopen{<}A, C, \tau\mathclose{>}$, it is enough to
provide the function $\tau$, which in its set-theoretical realization is
just a set of pairs. In relational programming, a set can be easily
encoded as a unary relation that answers the membership test. Therefore,
our goal will be to construct to answer this relation and this relation
will be our final representation of a signature.

We have our lexical database, a set of pairs of wordforms and hypertags,
and we want to arrive at a set of pairs of constants and their
types. This mapping decomposed into smaller parts by considering the
contribution of every part of the lexical database individually. What
will be the right model for the link between some part of the lexical
database and some part of our signature?

\begin{description}
  \item[Function of wordform] We could imagine that every different
    wordform of the lexical database will end up being mapped to some
    constant in the signature. However, this is unsuitable since the
    lexical database is a relation that can assign more than one
    distinct hypertag to a single wordform (e.g. with different part of
    speech) and we would like to have constants of different types for
    each of these hypertags.
  \item[Function of hypertag] Conversely, we might imagine that the
    hypertags of the lexical database are mapped via a function to the
    constants of the signature. Here we run into trouble when we
    consider variations in spelling or phonology. We might have two
    items in our lexical database that share the same hypertag but whose
    wordforms still differ. We would like to have two distinct constants
    in our signature then, so that when these two constants are mapped
    by the surface lexicon, they will yield two different strings.
  \item[Function of lexical entry] Since neither the wordform nor the
    hypertag is enough, we could consider the entire lexical entry (an
    element of the lexical database, a pair of a wordform and one of its
    hypertags). This solves both of the above problems but it is still
    not perfect. Consider the case of a determiner and some semantic
    signature. Since our link between elements of the lexical database
    and of our signature is a function, there must be a constant in the
    signature for every item of our lexical database. If you recall
    in~\ref{ssec:acg-sem}, the constants of the semantic signature
    $\Sigma_{Sem}$ contained predicates for the common nouns $man$ and
    $woman$ and for the verb $loves$ and some common logical
    furniture. However, it did not contain any constants which would
    correspond to the determiners $every$ and $some$.
  \item[Partial function of lexical entry] We could fix the issue above
    by using a partial function. This accounts for cases when a lexical
    entry has no constants to represent it in a signature. Nevertheless,
    this only fixes half of the problem. We will consider the single
    lexical entry which represents e.g. the 3rd person singular present
    tense form $loves$ of the transitive verb $love$. A legitimate
    account of subject/object scope ambiguity would be to consider two
    constants in our abstract syntactic signature that have distinct
    images in some semantic lexicon, one giving the outer scope to the
    subject and the other to the object. This means that in order to
    define such a signature, we would need to link a single lexical
    entry in our database to more than one constant in the signature.
  \item[Relation between lexical entry and constant] Finally,
    considering all the above common scenarios, the only mathematical
    model general enough to handle all of the above cases is that of a
    relation. In our system, this will correspond exactly to a relation
    that will be provided by the grammar author.
\end{description}

We have established that lexicons will be defined by a relation, a
relation between a lexical entry (a wordform and a hypertag) and a
constant (an identity\footnote{The identity of a constant is simply what
  distinguishes it from other constants of the same signature having the
  same type.} and a type). The identity of a lexical constant also has a
complex structure. We know that the lexical constants we define using
the relation described above are always linked to a lexical entry. This
lexical entry at least partially identifies the constant. For cases
where we have more than one constant per lexical entry (such as the
example of the two different scope readings of $loves$), we need some
extra piece of information to specify which constant we are referring to
(we call this extra piece of information a \emph{specifier}). This
specifier can have any shape that the grammar author desires. We can see
an example of a lexical constant in Figure~\ref{fig:lex-const}.

\begin{figure}
  % TODO: Center this.
  \centering
\begin{verbatim}
{:type [-> NP [-> NP S]]
 :id {:lex-entry {:wordform "loves"
                  :hypertag ...}
      :spec {:scope :subject-wide}}}
\end{verbatim}
  \caption{\label{fig:lex-const} An example of a lexical constant from a
    hypothetical syntactic signature (the hypertag is elided for brevity).}
\end{figure}

The system takes care of wiring up the relations and assembling the
signature. The grammar author's duty is to provide a 4-ary relation that
links the four parts of a lexical constant: the wordform, the hypertag,
the specifier and the type. This relation can be thought of as a
multi-valued ``function'' that receives as input a wordform and a
hypertag (that is, a single lexical entry) and outputs (possibly
multiple) pairs of specifiers and types. Since we are working in a
functional programming language that supports manipulation of
higher-order functions, we can provide helper functions which construct
these 4-ary relations for the grammar author in some simple cases. Our
library thus provides functions like \texttt{unitypedr}, which expects a
type and returns the 4-ary relation that assigns to every lexical item a
single constant whose specifier is $nil$ and whose type is the supplied
type, or \texttt{ht->typer}, which expects a mapping from hypertag
patterns to types and returns the 4-ary relation which tries to map the
hypertag against the patterns and yields constants with the
corresponding types and $nil$ specifiers. When other patterns emerge,
the grammar author is completely free to implement her new abstractions
using the full power of the functional programming language. See
Figure~\ref{fig:lex-sig-impl} for an example of implementing a
lexicalized version of the $\Sigma_{Synt}$ signature
from~\ref{ssec:example-sig} using the \texttt{ht->typer} abstraction.

\begin{figure}
  \centering
\begin{verbatim}
(ht->typer {{:head {:cat "n"}}      'N
            {:head {:cat "v"
                    :trans "true"}} (-> 'NP 'NP 'S)
            {:head {:cat "det"}}    (-> 'N (-> 'NP 'S) 'S)})
\end{verbatim}
  \caption{\label{fig:lex-sig-impl} A lexicalized version of the
    $\Sigma$(\_{Synt}) signature from~\ref{ssec:example-sig}.}
\end{figure}

In the above paragraphs, we have been referring to the constants that we
were defining as \emph{lexical constants}. That was to distinguish them
from the \emph{non-lexical constants} that we will introduce now. The
reason for having non-lexical constants is that not every constant of
every signature is linked to (generated by) some item of our lexical
database. Consider, for example, the empty string $\epsilon$ and the
string concatenation operator $+$ of the $\Sigma_{String}$ signature we
gave in~\ref{ssec:example-sig} or the quantifiers and logical
connectives of the $\Sigma_{Sem}$ signature in~\ref{ssec:acg-sem}.
These constants are not associated to any lexical entry or specifier and
therefore their representation differs. See
Figure~\ref{fig:nonlex-const} for an example of an non-lexicalized
constant from a semantic signature.

\begin{figure}
  \centering
\begin{verbatim}
{:type [-> T [-> T T]]
 :id {:constant-name and?}}
\end{verbatim}
  \caption{\label{fig:nonlex-const} An example of the representation of
    a non-lexicalized constant (the conjunction operator in a semantic
    signature).}
\end{figure}

Signatures of non-lexical constants can be simply defined by a mapping
from the constant names to their types as
in~\ref{fig:nonlex-sig-impl}. Such signatures will generally not be
sufficient by themselves and will need to be complemented by other
signatures which contain the lexical constants. Since we represent
signatures as unary relations, combining them to produce their unions or
intersections is just a matter of taking their disjunctions or
conjunctions, respectively. With this, we have enough to fully define
the other signatures of section~\ref{sec:acg} in a lexicalized
manner, see Figure~\ref{fig:example-sig-impl}.

\begin{figure}
  \centering
\begin{verbatim}
(nonlex-sigr {'and?    (-> 'T 'T 'T)
              'imp?    (-> 'T 'T 'T)
              'forall? (-> (=> 'E 'T) 'T)
              'exists? (-> (=> 'E 'T) 'T)})
\end{verbatim}
  \caption{\label{fig:nonlex-sig-impl} A signature of non-lexical
    constants belonging to $\Sigma$(\_{Sem})
    from~\ref{ssec:acg-sem}. The single arrow \texttt{->} corresponds
    to linear implication while the double arrow \texttt{=>} corresponds
    to intuitionistic implication.}
\end{figure}

@@

(def sem-sig
  (ors (nonlex-sigr {'and?    (-> 'T 'T 'T)
                     'imp?    (-> 'T 'T 'T)
                     'forall? (-> (=> 'E 'T) 'T)
                     'exists? (-> (=> 'E 'T) 'T)})
       (ht->typer {{:head {:cat "n"}}      (-> 'E 'T)
                   {:head {:cat "v"
                           :trans "true"}} (-> 'E 'E 'T)})))
\end{verbatim}
  \caption{\label{fig:example-sig-impl} The definition of the
    lexicalized versions of the example signatures of
    section~\ref{sec:acg}.}
\end{figure}

\subsection{Defining lexicons}

We have explained how we can use elements of our lexical database to
define signatures. Now, we will turn to lexicons. Similar to how we
reduced signatures from the formal triples $\mathopen{<}A, C,
\tau\mathclose{>}$ to just the type assignment function $\tau$, we can
similarly reduce lexicons from the pairs $\mathopen{<}F, G\mathclose{>}$
to only $G$, the part which operates on terms. $F$, or at least its
relevant subset, can be easily retrieved by putting the types of the
abstract constants side-by-side with the types of their object terms to
get a sufficient subset of $\hat{F}$ and finding $F$, the mapping from
atomic abstract types whose homomorphic extension includes $\hat{F}$.

In our relational framework, a mapping is most directly represented as a
binary relation, which is the representation that we have chosen. In the
previous subsection, we talked a lot about how individual entries in
the lexical database generate the items of a signature. Solving the
problem for signatures also solved it for lexicons, since their
structure is very similar: a signature is a set of pairs of constants
and types, a lexicon is a set of pairs of constants and
terms. Therefore, a lexical entry contributes the object terms for the
constants which it generated in the abstract signature.

Our representation of lexical constants was purposefully engineered so
as to make the lexical entry directly accessible to the lexicon
implementation. This way, the lexicon author can inspect the hypertag
and the wordform of the lexical entry, the specifier of the concrete
constant that it is mapping right now and even the type she assigned to
the constant in the abstract signature to allow her to choose the
correct object term. Furthermore, since lexicons are still
relations/sets, we can compose them together by disjunction. See
Figure~\ref{fig:lex-impl} for the implementation of the lexicalized
versions of the $\mathcal{L}_{Syntax}$ and $\mathcal{L}_{Sem}$ lexicons.

@@

(def sem-lexo
  (with-sig-consts sem-sig
    (orr (lexicalizer sem-sig
                      (ht-lexiconr {{:head {:cat "n"}}
                                    ,(rt (ll [_ x] (_ x)))
                                    {:head {:cat "v"
                                            :trans "true"}}
                                    ,(rt (ll [_ s o] (_ s o)))}))
         (ht-lexiconr {{:head {:cat "det"
                               :lemma "un"}}
                       ,(rt (ll [P Q] (exists? (il [x] (and? (P x)
                                                             (Q x))))))
                       {:head {:cat "det"
                               :lemma "chaque"}}
                       ,(rt (ll [P Q] (forall? (il [x] (imp? (P x)
                                                             (Q x))))))}))))
\end{verbatim}
  \caption{\label{fig:lex-impl} The implementation of the
    $\mathcal{L}$(\_{Syntax}) and $\mathcal{L}$(\_{Sem}) lexicons from
    section~\ref{sec:acg}.}
\end{figure}

 NOTE: This is too verbose both to put on the paper and also to
 code. These examples demonstrate two common and related shapes of
 lexicons. The $\mathcal{L}$(\_{Syntax}) and the first half of the
 $\mathcal{L}$(\_{Sem}) are lexicalized, in the sense that the object
 term contains a lexical constant with the same lexical entry as the
 abstract constant.\footnote{This notion of ``lexicalized'' is quite
   different to the one I use in the rest of the thesis so far (which
   roughly means ``generated by a lexical database'') and also different
   from the one used by Kanazawa (FIX!).} This means that before
 constructing the object term, we have to acquire the object constant
 that is contained in the term and then we just fit it in of the
 recipes, which we choose according to the hypertag. The second half of
 $\mathcal{L}$(\_{Sem}) simply assigns object terms based on the
 hypertag. A more concise and desirable implementation would allow us
 simply write the pairs of hypertags and object terms (recipes encoded
 as lambdas). My plan is to have exactly this, but I have not thought
 yet about how to split the various aspects into different reusable
 abstractions.

\section{Testing signatures and lexicons}

>>>>

\chapter{Implementing Wide-Coverage Abstract Categorial Grammars}
\label{chap:implementation}

@@

Given this setup, there are several approaches we might try, none of
them too appealing. We might write our metagrammar in a different format
and use a tool which would combine the metagrammar and the lexical
database\footnote{We will be using the term \emph{lexical database} to
  refer to what is usually called a lexicon, in order not to cause
  confusion with the lexicons of ACGs.} to produce the
lexicalized\footnote{We will be using the term \emph{lexicalized} in the
  context of ACGs to mean that the elements of the grammar (the typed
  constants and their images given by lexicons) are generated by a
  lexical database. This is not to be confused with the term used in
  \cite{yoshinaka2005complexity}.} ACG in a direct representation, ready
to be loaded by the toolkit. This workflow would make development on the
grammar tedious as every time we would change the metagrammar, we would
have to regenerate the direct representation and then load it in the ACG
toolkit. With extra effort, the former cost could be alleviated by
devising a system for regenerating only parts of the grammars which will
have changed since last time. However, we would still have to pay the
price of loading the entire grammar into the toolkit before being able
to interact with it again.

@@

In this chapter, we will spend some time explaining the design rationale
underlying the system and the form it has now. The source code of the
system in its current state can be perused at
\url{https://github.com/jirkamarsik/acg-clj}.

\section{The Tools and Techniques}

@@

By insisting on purity, I/O becomes more difficult as we would not allow
the side-effectful pseudo-relations of Prolog. Therefore, we will not
force ourselves to implement the entire program front-to-back using only
relational programming. Instead, we will use a relational programming
system, miniKanren, which is embedded inside a functional programming
language. This lets us reap the benefits of relational programming
inside the core logic of our program while deferring to the functional
programming language to provide I/O and other bookkeeping operations.

\subsection{The Case for Relational Programming}

There are several reasons why we might want to choose a relational
programming system for our implementation. First off, the problems which
constitute the algorithmic core of our program are often mere textbook
examples of how to use a relational programming system. This goes for
type inference/checking/inhabitation, which are all implemented by a
single relation which is often used as a kind of ``Hello World'' program
in the world of relational programming. Similarly, implementing
$\beta$-reduction cleanly (without having to handle variable clashes by
generating new variables and $\alpha$-converting) is a motivational
example for introducing \emph{nominal logic programming}. To reach
feature parity with the ACG toolkit, one would only need to implement
the mapping of a lexicon over a term, which is a simple functor.

@@

Finally, some of the structures that we will encounter in our domain are
more directly modelled in mathematics as relations rather than
functions. In a relational programming system, to implement a relation,
one just writes a relation as it is, whereas if we were forced to encode
everything in functions, we would have to simulate relational behavior
manually. Frilex, our lexical database, is an example of a structure
that has this relational shape since a single wordform can be mapped to
multiple hypertags. We will see another example of a relation within our
domain in \ref{ssec:def-sig} and we will also discuss the advantages and
drawbacks of modelling functions using relations.

\subsection{Choice of Implementation Language}

There are two programming languages which play host to a substantial
implementation of miniKanren (meaning one that has extensible
unification and/or constraints and that includes support for nominal
logic), Racket\footnote{\url{https://github.com/calvis/cKanren}} and
Clojure\footnote{\url{https://github.com/clojure/core.logic}}. The two
languages are very similar to each other and therefore the choice is of
no great significance. An appealing feature of Clojure is that it uses
abstract collection types pervasively instead of coupling the standard
operations to specific data structures such as lists and cons cells and
that it follows other similarly enlightened design decisions. Racket on
the other hand has had more time to mature, has better metaprogramming
facilities and is more recognized in academia. In developing our system,
we have settled for Clojure. One of the advantages of doing so is that
the Clojure implementation of miniKanren, dubbed core.logic, has
inspired the software community and the project has attracted
contributors that submit numerous issue reports and
patches.\footnote{\url{http://dev.clojure.org/jira/browse/LOGIC}}

\section{Defining Signatures and Lexicons}

In this section, we will go over the representation of signatures and
lexicons in our implementation and the basic facilities for defining
them.

\subsection{Defining Signatures}
\label{ssec:def-sig}

In~\ref{ssec:sig}, we formally presented signatures as triples
$\mathopen{<}A, C, \tau\mathclose{>}$. The set $C$ can be easily
reconstructed as the domain of $\tau$ and the set $A$ is just the set of
all atomic types occurring in the range of $\tau$. Therefore, to define
the signature $\mathopen{<}A, C, \tau\mathclose{>}$, it is enough to
provide the function $\tau$, which in its set-theoretical realization is
just a set of pairs. In relational programming, a specific set can be
easily encoded as a unary relation so a unary will serve as our
representation of a signature and defining a signature will be a matter
of constructing this relation.

\subsubsection{The Relationship Between the Lexical Database and a Signature}

We have our lexical database, a set of pairs of wordforms and hypertags,
and we want to arrive at a set of pairs of constants and their
types. This mapping can be decomposed into smaller parts by considering
the contribution of every part of the lexical database
individually. What will be the right model for the link between some
part of the lexical database and some part of our signature?

\begin{description}
  \item[Function of wordform] We could imagine that every different
    wordform of the lexical database will end up being mapped to some
    constant in the signature. However, this is unsuitable since the
    lexical database is a relation that can assign more than one
    distinct hypertag to a single wordform (e.g. a wordform with
    different possible part of speech) and we would like to have
    constants of different types for each of these hypertags.
  \item[Function of hypertag] Conversely, we might imagine that the
    hypertags of the lexical database are mapped via some function to
    the constants of the signature. Here we run into trouble when we
    consider variations in spelling or phonology. We might have two
    items in our lexical database that share the same hypertag but whose
    wordforms still differ. We would then like to have two distinct
    constants in our signature as well, so that when these two constants
    are mapped by the surface lexicon, they will yield two different
    strings.
  \item[Function of lexical entry] Since neither the wordform nor the
    hypertag is enough, we could consider the entire lexical entry (an
    element of the lexical database, a pair of a wordform and one of its
    hypertags). This solves both of the above problems but it is still
    not perfect. Consider the case of a determiner and some semantic
    signature. Since our link between elements of the lexical database
    and of our signature is a function, there must be a constant in the
    signature for every item of our lexical database. If you recall
    in~\ref{ssec:acg-sem}, the constants of the semantic signature
    $\Sigma_{Sem}$ contained predicates for the common nouns \emph{man}
    and \emph{woman} and for the verb \emph{loves} and some common
    logical furniture. However, it did not contain any constants which
    would correspond to the determiners \emph{every} and \emph{some} and
    therefore there would be no constant that the function could assign
    to these two wordforms.
  \item[Partial function of lexical entry] We could fix the issue above
    by using a partial function. This accounts for cases when a lexical
    entry has no constants to represent it in a signature. Nevertheless,
    this only fixes half of the problem. We will consider the single
    lexical entry which represents e.g. the 3rd person singular present
    tense form \emph{loves} of the transitive verb \emph{love}. A
    legitimate account of subject/object scope ambiguity would be to
    consider two constants in our abstract syntactic signature that have
    distinct images in some semantic lexicon, one giving the outer scope
    to the subject and the other to the object. This means that in order
    to define such a signature, we would need to link a single lexical
    entry in our database to more than one constant in the signature.
  \item[Relation between lexical entry and constant] Finally,
    considering all the above scenarios, the only mathematical model
    general enough to handle all of these cases is a relation. In our
    system, this will correspond exactly to a specific relation that
    will be provided by the grammar author.
\end{description}

\subsubsection{Representing Constants and Defining Signatures}

We have established that signatures will be defined by a relation, a
relation between a lexical entry (a wordform and a hypertag) and a
constant (an identity\footnote{The identity of a constant is simply the
  thing that distinguishes it from other constants of the same signature
  having the same type.} and a type). The identity of a lexical constant
also has a complex structure. We know that the lexical constants we
define using the relation described above are always linked to a lexical
entry. This lexical entry at least partially identifies the
constant. For cases where we have more than one constant per lexical
entry, such as in the example of the two different scope readings of
\emph{loves}, we need some extra piece of information to specify exactly
which of the two constants we are referring to (we call this extra piece
of information a \emph{specifier}). This specifier can have any shape
that the grammar author desires. We can see an example of a lexical
constant in Figure~\ref{fig:lex-const}.

\begin{figure}
  \centering
\begin{verbatim}
{:type [-> NP [-> NP S]]
 :id {:lex-entry {:wordform "loves"
                  :hypertag ...}
      :spec {:scope :subject-wide}}}
\end{verbatim}
  \caption{\label{fig:lex-const} An example of a lexical constant from a
    hypothetical syntactic signature (the hypertag is elided for brevity).}
\end{figure}

The system takes care of wiring up the relations and assembling the
signature. The grammar author's duty is to provide a 4-ary relation that
links the four parts of a lexical constant: the wordform, the hypertag,
the specifier and the type. This relation can be thought of as a
multi-valued ``function'' that receives as input a wordform and a
hypertag (that is, a single lexical entry) and outputs (possibly
multiple) pairs of specifiers and types.

Since we are working in a functional programming language that supports
manipulation of higher-order functions, we can provide helper functions
which construct these 4-ary relations for the grammar author in some
simple cases. Our library thus provides functions like
\texttt{unitypedr}, which expects a type and returns the 4-ary relation
that assigns to every lexical item a single constant whose specifier is
$nil$ and whose type is the supplied type, or \texttt{ht->typer}, which
expects a mapping from hypertag patterns to types and returns the 4-ary
relation which tries to map the hypertag against the patterns and yields
constants with the corresponding types and $nil$ specifiers. When other
patterns emerge, the grammar author is completely free to implement her
new abstractions using the full power of the functional programming
language.

See Figure~\ref{fig:lex-sig-impl} for an example of implementing a
lexicalized version of the $\Sigma_{Synt}$ signature
from~\ref{ssec:example-sig} using the \texttt{ht->typer} abstraction.

\begin{figure}
  \centering
\begin{verbatim}
(ht->typer {{:head {:cat "n"}}      'N
            {:head {:cat "v"
                    :trans "true"}} (-> 'NP 'NP 'S)
            {:head {:cat "det"}}    (-> 'N (-> 'NP 'S) 'S)})
\end{verbatim}
  \caption{\label{fig:lex-sig-impl} A lexicalized version of the
    syntactic signature from~\ref{ssec:example-sig}.}
\end{figure}

In the above paragraphs, we have been referring to the constants that we
were defining as \emph{lexical constants}. That was to distinguish them
from the \emph{non-lexical constants} that we will introduce now. The
reason for having non-lexical constants is that not every constant of
every signature is linked to (generated by) some item of our lexical
database. Consider, for example, the empty string $\epsilon$ and the
string concatenation operator $+$ of the $\Sigma_{String}$ signature we
gave in~\ref{ssec:example-sig} or the quantifiers and logical
connectives of the $\Sigma_{Sem}$ signature in~\ref{ssec:acg-sem}.
These constants are not associated to any lexical entry or specifier and
therefore their representations must be different. See
Figure~\ref{fig:nonlex-const} for an example of the representation we
use for non-lexicalized constants.

\begin{figure}
  \centering
\begin{verbatim}
{:type [-> T [-> T T]]
 :id {:constant-name and?}}
\end{verbatim}
  \caption{\label{fig:nonlex-const} An example of the representation of
    a non-lexicalized constant (in this case, it is the conjunction
    operator in a semantic signature).}
\end{figure}

Signatures of non-lexical constants can be simply defined by a mapping
from the constant names to their types as
in~\ref{fig:nonlex-sig-impl}. Such signatures will generally not be
sufficient by themselves and will need to be complemented by other
signatures which contain lexical constants. Since we represent
signatures as unary relations, combining them to produce their unions or
intersections is just a matter of taking their disjunctions
(\texttt{ors}) or conjunctions (\texttt{ands}), respectively.

\begin{figure}
  \centering
\begin{verbatim}
(nonlex-sigr {'and?    (-> 'T 'T 'T)
              'imp?    (-> 'T 'T 'T)
              'forall? (-> (=> 'E 'T) 'T)
              'exists? (-> (=> 'E 'T) 'T)})
\end{verbatim}
  \caption{\label{fig:nonlex-sig-impl} A signature of the non-lexical
    semantic constants belonging to the semantic signature
    of~\ref{ssec:acg-sem}. The single arrow \texttt{->} corresponds to
    linear implication while the double arrow \texttt{=>} corresponds to
    intuitionistic implication.}
\end{figure}

With this, we have enough to fully define all the signatures of
section~\ref{sec:acg} in a lexicalized manner, see
Figure~\ref{fig:example-sig-impl}.

@@

(def sem-sig
  (ors (nonlex-sigr {'and?    (-> 'T 'T 'T)
                     'imp?    (-> 'T 'T 'T)
                     'forall? (-> (=> 'E 'T) 'T)
                     'exists? (-> (=> 'E 'T) 'T)})
       (ht->typer {{:head {:cat "n"}}      (-> 'E 'T)
                   {:head {:cat "v"
                           :trans "true"}} (-> 'E 'E 'T)})))
\end{verbatim}
  \caption{\label{fig:example-sig-impl} The definitions of the
    lexicalized versions of the example signatures of
    section~\ref{sec:acg}.}
\end{figure}

\subsection{Defining Lexicons}

We have explained how we can use elements of our lexical database to
define signatures. Now, we will turn to lexicons. Similar to how we
reduced signatures from the formal triples $\mathopen{<}A, C,
\tau\mathclose{>}$ to just the type assignment function $\tau$, we can
reduce lexicons from the pairs $\mathopen{<}F, G\mathclose{>}$ to just
$G$, the part which operates on terms. $F$, or at least its relevant
subset, can be easily retrieved from $G$. We take the pairs of abstract
constants and object terms that $G$ is composed of, we take their types
and we arrive at pairs of abstract-level and object-level types. These
pairs form a subset of $\hat{F}$. $F$ (or at least its relevant subset),
which only maps the atomic abstract types, can be inferred from this
subset of $\hat{F}$.

In our relational framework, a mapping can be represented most directly
as a binary relation and that is the representation that we use in our
system. In the previous subsection, we talked a lot about how individual
entries in the lexical database generate the items of a
signature. Solving the problem for signatures also solved it for
lexicons, since a lexicon contains exactly one object term for every
constant in its abstract signature. Therefore, a pair of an abstract
constant and an object term belonging to some lexicon is generated by
the same lexical entry that generated the abstract constant.

Our representation of lexical constants was purposefully engineered so
as to make the lexical entry directly accessible to the lexicon
implementation. Since the lexicon is directly implemented by the grammar
author as a binary relation between constants and terms, the grammar
author can inspect the abstract constant and directly access the
hypertag and wordform of the lexical entry that generated it, the
specifier that was assigned to it and the type of the constant and then
express the object term assigned to the constant using any combination
of these.

\subsubsection{Succinct Definitions of Lexicons}

The above is already enough to define any lexicon. However, our toolkit
presents convenience facilities that capture common forms of
lexicons. With these, we will be able to define lexicalized versions of
the lexicons of section~\ref{sec:acg} in a concise point-free
style. Furthermore, the grammar author is not limited to the convenience
facilities we have provided and is completely free to use the same
programming language we have used to build on our abstractions and build
her own right within her grammar definition.

@@

(def sem-lexo
  (with-sig-consts sem-sig
    (orr (lexicalizer sem-sig
                      (ht-lexiconr {{:head {:cat "n"}}
                                    ,(rt (ll [_ x] (_ x)))
                                    {:head {:cat "v"
                                            :trans "true"}}
                                    ,(rt (ll [_ s o] (_ s o)))}))
         (ht-lexiconr {{:head {:cat "det"
                               :lemma "un"}}
                       ,(rt (ll [P Q] (exists? (il [x] (and? (P x)
                                                             (Q x))))))
                       {:head {:cat "det"
                               :lemma "chaque"}}
                       ,(rt (ll [P Q] (forall? (il [x] (imp? (P x)
                                                             (Q x))))))}))))
\end{verbatim}
  \caption{\label{fig:lex-impl} The definitions of the lexicalized
    versions of the lexicons of section~\ref{sec:acg}.}
\end{figure}

We have the lexicon analogues to \texttt{nonlex-sigr} and
\texttt{ht->typer}, they are called \texttt{nonlex-lexiconr} and
\texttt{ht-lexiconr} and they work the same way as the signature
versions but assign object terms instead of types.

When mapping lexical constants by a lexicon, we usually produce object
terms which contain constants of the object signature that are generated
by the same lexical entry as the abstract constant
(e.g. $\mathcal{L}_{syntax}(\synt{love}) = \lambda^{\circ} x y.\ x +
loves + y$). Managing this involves some tedious boilerplate that can be
avoided by another convenience facility. First, we define a lexicon that
produces the desired object term but with the object constant abstracted
out (e.g. $\lambda^{\circ} c x y.\ x + c + y$). Then, we apply the
\texttt{lexicalizer} function to the object signature and to this
lexicon to get the desired lexicon which fills in the correct object
constants.

It is also useful to be able to easily insert the non-lexical constants
of the object signature in the object terms we are giving when defining
a lexicon. This service is provided by the anaphoric macro
\texttt{with-sig-consts} which introduces all of the non-lexical
constants of a signature into scope for easy use inside of lexicon
definitions.

Finally, since lexicons are again just sets of pairs, we can take their
unions and intersections using disjunction (\texttt{orr}) and
conjunction (\texttt{andr}).

This gives us everything we need to define the lexicalized versions of
the lexicons of section~\ref{sec:acg} in a concise and elegant
manner. See Figure~\ref{fig:lex-impl} for the definitions.

\section{Testing Signatures and Lexicons}
